---
title       : A gentle introduction to R
subtitle    : for Stata user
author      : C.V. An
job         : an avid R user
framework   : io2012   # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
github:
  user: anchuvan
  repo: Introductory R
---

## Table of contents

1. What is R
2. Programming with R
3. Statistics with R
4. Graphics with R

--- .class #id 

## 1. What is R

>* A statistical package (competitors: Stata, SAS, SPSS, Matlab...)

>* A programming language (object-oriented) 

>* An environment (a huge system of algorithms, packages)

>* A community (over 2 million users worldwide)

>* A de-facto standard among statisticians for the development of statistical methods
	and teaching

---

## Why R

>* Free + Open source
>* Flexible
	  * Open for integration: 
		  * Data: SAS, SPSS, Excel, STATA, SQL Server,...
		  * Systems: applications, webservers,...
	  * Able to combine with other programming languages (Python, C, C++,...)
>* Well-supported community (stackoverflow, github, mailing list, r-blogs,...)
>* Excellent graphics (the best among statistical packages)
>* Highly extensible (over 4300 user-contributed packages)
>* General purpose (statistical analysis, data visualization, machine learning,...)
>* Reproducible research(R + LaTeX)

---

## Install R

* Go to [http://r-project.org](http://www.r-project.org)
* Select any mirror (Vietnam)
* Choose your platform (Windows) and click <i>install R for the first time</i>
* Download <b>R 3.0.0.0 for Windows</b>

---

## Get to know Rgui

__Command line interface (CLI)__: R Console -> an interactive enviroment (workspace)
```{r}
1 + 1
x <- "Hello Stata user"
print(x)
```

__R script editor__ (similar to Stata do-file)

---

## Find help

__Access help file__
```{r,tidy=FALSE, eval=FALSE}
?rnorm    ## returns documentation of 'rnorm'
```
__Search help files__
```{r, eval=FALSE, echo=T}
help.search("rnorm")  ## finding all functions having keyword: 'rnorm'
## equivalent to: ??rnorm
```
__Get arguments__
```{r, eval=FALSE, echo=TRUE}
args("rnorm")  ## outputs all parameters used by rnorm
````
__Example__
```{r, eval=F, echo=TRUE}
example("mean") ## give some examples of how to use 'mean'
```
---

## Set your working directory

* All R files or datasets are recommended to store in a defined directory

* Use command: `getwd()` to get your current working directory

* Use command: `setwd()` to change your working directory

* For exampple: `setwd("C:/user/antom/Documents/R")` 

---

## 2. Programming with R

* 2.1 How R works
* 2.2 Object-oriented programming
* 2.3 Data type
* 2.4 Data structure
* 2.5 Getting data in

---

## 2.1 How R works

>* R deploys packages to work on specific tasks. Each package comes with a wide
   range of functions, methods, and datasets.

>* CRAN TASK VIEWS classify all packages in different topics
   [http://cran.r-project.org/web/views/](http://cran.r-project.org/web/views/)

>* Default packages in R. 
   * type: `sessionInfo()` to see loaded packages on workspace
   * type: `installed.packages()` to see installed packages in your computer

>* What's inside an package? 
   * type: `help(package = datasets)` for full detail

>* How to install packages?
   * type: `install.packages("package_name")`, then select any mirror (usually, Vietnam)
   * after that, `library(package_name)` makes it available for use

---

## 2.2 About object-oriented programming

* Everything in R is an object

* Each object belongs to a class (or an instance of the class, 
   by which it inherits all chracteristics of the class)

* Class has its own <b>functions</b> which manipualte objects
   and <b>methods</b> that can be deployed by objects
                
* An example:
```{r6, tidy=F, eval=F, echo=T}
x <- 123; y <- 246 ## creat two objects x and y
sum(x + y) ## function: sum() calculates the sum value of two objects x and y
as.character(x) ## function: as.character() transform data type of object x
#######
y - x    ## method: substraction, - (y x)
```
---

## 2.3 Basic data type in R

* Integer: `1, 2, 5`
* Numeric: `1.23`
* Logical: `TRUE, FALSE (abbr: T, F)`
* Complex: `1 + 3i`
* Character (or string): `"character" (or 'character')`
* Factors: `Yes or No`

---

## 2.4 Basic data structure in R

* Vector
* List 
* Matrix
* Array
* Data.frame

---

## 2.4.1 Working with vector

* Vector: the atomic data structure in R, all elements of vector
  must be the same data type

* Declaration

* Common vector operations
  * Arithemetic and logocal operations
  * Vector indexing
  * Generating vector with `:` and `seq()` and `rep()`

* Vectorized operations

---

## Working with vector: declaration

__Using c()__
```{r, eval=FALSE}
x <- c(1, 2, 3)
y <- c("A", "b", "C", "d")
z <- c(TRUE, FALSE, TRUE)
g <- c(1 + 3i, 2 - 5i)
```
__Using vector()__
```{r, eval=F, echo=TRUE, tidy=FALSE}
y <- vector() ## return a vector with length = 0
y <- vector(length = k) ## return a vector length = k
```
---

## Working with vector: Arithemetic and logical operations
```{r, tidy=F, eval=T, echo=TRUE}
c(1, 2, 3) + c(2, 4, 6)
c(2, 4, 6)/c(4, 6, 8)
c(c(1,2,3), c(5, 6))  # 
```
---

## Working with vector: indexing
* Problem: want to extract individual/set of elements from a vector
```{r7, tidy=FALSE, echo=TRUE, eval=FALSE}
age <- c(22, 23, 24, 26, 36, 36, 36, 23)  ## create a vector
age[1]  ## return the first element; age[k] return the k element
#[1] 22
age[1:3] ## return the three fist elements
#[1] 22 23 24
age[c=(1, 2, 3)] ## same output as age[1:3]
#[1] 22 23 24
age[-3]  ## return all elements except for the third element
#[1] 22 23 26 36 36 36 23
age[3, -2] ## Error, can't index both positive and negative value
```
---

## Working with vector: vectorization
```{r8, tidy=F, echo=T, eval=T}
x <- c(12, 34, 54); y <- c(28, 98, 56)
x + y
x/y
x * 3
```
---

## Working with vector: generating vector
```{r, tidy=F, eval=T, echo=T}
1:5   # seq(1:5) 
seq(1,5, by=2) # seq(from= , to= , by= )
rep(1:3, times=4) ## rep(x= , times=, each= )
```
---

## 2.4.2 Working with list

* Basically, list is a nested vector.
* Example: `my.list <- list(a=c(1:10), b=c("this", "is", "it"))`
```{r}
my.list <- list(a=c(1:10), b=c("this", "is", "it"))
my.list ## this list contains 2 elements, each elements is a vector
```
---

## Working with list: indexing
```{r9, tidy=F, echo=T, eval=T}
my.list <- list(a=c(1:10), b=c("this", "is", "it"))
my.list[[1]]  ## indexing by number
names(my.list)
my.list["a"]  ## indexing by name
```
---

## 2.4.3 Working with data frame

* Data frame is a 2x2 matrix, but it can handle different type of data

* Data frame is a rectangular table with columns represent 
  variables and rows represent obsevation

* How to create a data frame: ` df <- data.frame(c1, c2)` 
  with c1 and c2 are vectors

---

## 2.5 Getting data in

* Reading data directly from CLI by `scan()`

* Using `library(foreign)`
* Make sure your data file in your current working directory
  * `getwd()` returns your current working directory
  * `setwd` setting your working directory, e.g: `setwd("C:/user/R")`
* Basic syntax:
  * data <- read.dta("filename.dta",...)
  * data <- read.spss("filename.sav",...)
  * data <- read.csv("filenam.csv",...)
  * data <- read.table("filename.txt",...)
  
* `?read.table` for more instructions  

---

## 3. Statistics with R

* 3.1 Desciptive statistics

* 3.2 Inferential statistics

---

## 3.1 Descriptive statistics

__Data set__
```{r, tidy=FALSE}
set.seed(10) ## ensure the sample can be reproducible
x <- sample(x = 1:100, size = 20, replace = TRUE) ## choose a random sample from vector x
print(x)  
```
---

## 3.1 Desciptive statistics

__Descriptive statistics__
```{r, results='hide', tidy=F}
min(x) ## [1] 7
max(x) ## [1] 100
mean(x) ## [1] 56
median(x) ## [1] 60.5
range(x) ## 
summary(x) ## similiar to summarize in Stata
```
---

## 3.2 Inferential statistics (regression model)

__Reading in data__
```{r}
data(cars)
str(cars)
```
---

## 3.2 Infrential statistics (regression model)

__Linear regression model:__
* R use function lm() to analyze regression model, all the results can be stored in an object,
  and this object (regress) is a <b>list</b>
```{r, eval=F, echo=T}
regress <- lm(dist ~ speed, data=cars)
summary(regress) ## to print out all results
```

---

## 3.2 Infrential statistics (regression model)

__Linear regression model:__
```{r, eval=T, echo=F}
regress <- lm(dist ~ speed, data=cars)
summary(regress)
```
---

## 3.2 Infrential statistics (regression model)
__regress is a object__
```{r, eval=T, echo=T}
regress <- lm(dist ~ speed, data=cars)
names(regress)
```
---

## 3.2 Infrential statistics (regression model)
```{r}
regress$residuals
```
---

## 3.2 Infrential statistics (regression model)
```{r}
boxplot(regress$residuals)
```
---

## 3. Graphics with R

* R provides the usual range of standard statistical plots, including scatterplots,
boxplots, histograms, barplots, piecharts, and basic 3D plots

* Basically, these plots can be produced by a single function call. (high-level functions)

* R is also able to add several graphical elements together. (text, lines, points, axis,...)

---

## First example of R graphics
```{r, tidy=F, fig.align='center', eval=TRUE, echo=TRUE}
plot(pressure)  # high-level funtion
text(150, 600, "Hello Stata Users!") # adding graphical element: text
```

## High-level functions:

* Scatter plot: `plot()`, can be used for dot plot and line chart
 
* Box and whisker plot: `boxplot()`

* Histogram: `hist()`

* Bar chart: `barplot()`

---

## High-level functions:
__ An example__
```{r, echo=T, eval=F, tidy=FALSE}
par(mfrow=c(2, 2)) ## create 4 panels
N <- 200
x <- runif(N, -4, 4)
y <- sin(x) + 0.5*rnorm(N)
plot(x, y, main="Scatter plot of y and x")
hist(x, main="Histogram of x")
boxplot(y, main="Box plot of y")
barplot(x, main="Bar chart of x")
```
---

## High-level functions
```{r, echo=F, eval=T, tidy=FALSE}
par(mfrow=c(2, 2)) ## create 4 panels
N <- 200
x <- runif(N, -4, 4)
y <- sin(x) + 0.5*rnorm(N)
plot(x, y, main="Scatter plot of y and x")
hist(x, main="Histogram of x")
boxplot(y, main="Box plot of y")
barplot(x, main="Bar chart of x")
```
---

## Scatter plot:
* Main parameters: `x=, y=, xlab=, ylab=, pch=, main=, type=, col=`

```{r, eval=T, echo=T, tidy=FALSE, fig.height=6, fig.width=6}
plot(x = cars$speed, y = cars$dist) 
```
---

## Scatter plot:
* R control characteristics of graphs via parameters: x, y, xlab, ylab,....

```{r, eval=F, echo=T, fig.align='center', tidy=FALSE}
plot(x=cars$speed, y=cars$dist,
       xlab = "Speed (miles per hour)", # rename x label
       ylab = "Stopping distance", # rename y label
       main = "Scatter plot",  # title of the plot
       cex.main = 1.6, ## Resize the title text
       tck = -0.015, ## Resize the tick mark
       pch = 16, col = "purple", ## choose another point character and set color
       las = 1, bty = "l",  # change the frame from "o" to "l"
	      col.lab = "red4", cex.lab = 1.3, # change color and size of label
	      col.axis = "black",  # change color of 
	      col.main = "red4") # change color of title
```
---

## Scatter plot:

```{r, eval=T, echo=F, fig.align='center', tidy=FALSE}
plot(x=cars$speed, y=cars$dist,
       xlab = "Speed (miles per hour)",
       ylab = "Stopping distance",
       main = "Scatter plot",
       cex.main = 1.6, ## Resize the title text
       tck = -0.015, ## Resize the tick mark
       pch = 16, col = "purple", ## choose another point character and set color
       las = 1, bty = "l",
        col.lab = "red4", cex.lab = 1.3,
	      col.axis = "black",
	      col.main = "red4")
```
---

## Scatter plot:

```{r, tidy=F, echo=T, fig.height=5, fig.width=9}
par(mfrow=c(1, 2))
plot(x=cars$speed, y=cars$dist)
plot(x=cars$speed, y=cars$dist, pch=16, col="purple", bty="l", las=1,
     xlab="Speed (mph)", ylab="Stopping distance", main="Scatter plot")
```
---

## Line chart
__Argument: type="l"__
```{r, tidy=F, echo=T, fig.height=5, fig.width=9}
par(mfrow=c(1, 2)); library(ggplot2)
plot(pressure$temperature, pressure$pressure, type="l")
plot(pressure$temperature, pressure$pressure, type="b")
```
---

## line charts
```{r, tidy=FALSE, echo=FALSE}
par(mfrow=c(2,2))
plot(y, type="l"); title("lines")
plot(y, type="b"); title("both")
plot(y, type="o"); title("overstruc")
plot(y, type="h"); title("high density")
```
---

## Line graph: 

```{r, eval=T, echo=F}
rain <- read.csv("C:/Users/antom/Documents/R/datasetc1/cityrain.csv", 
			header=TRUE)
plot(rain$Tokyo, type="l", col="red",
     ylim=c(0,250),
     main="Monthly rainfall in major cities",
     xlab="Month of Year",
     ylab="Rainfall (mm)",
     lwd=2,
     cex.main=1.3,
     las = 1, bty = "l",
     tck = -0.012)

lines(rain$NewYork,type="l",col="blue",lwd=2)
lines(rain$London,type="l",col="green",lwd=2)
lines(rain$Berlin,type="l",col="orange",lwd=2)

legend("top",
	 legend=c("Tokyo","NewYork","London","Berlin"),
	 ncol=4, ## number of legend columns (default is 1)
	 cex=0.8, ## size of legend
	 bty="n", ## type of border
	 col=c("red","blue","green","orange"),
	 lty=1,lwd=2)
```
---

## Box and whisker plot:
```{r, echo=T, tidy=F, eval=T, fig.align="centre"}
boxplot(len ~ supp + dose, data=ToothGrowth)
```

---

## Histogram:
```{r, tidy=F, echo=T, eval=T, fig.width=9, fig.height=5}
par(mfrow=c(1,2))
hist(mtcars$mpg)
hist(mtcars$mpg, breaks=10, col="steelblue")
```
---

## Bar chart:
```{r, echo=T, eval=T, tidy=F, fig.align='center'}
barplot(table(mtcars$cyl))
```
---

## Bar chart:
```{r, eval = T, echo =F, tidy=F}
citysales <- read.csv("C:/Users/antom/Documents/R/datasetc1/citysales.csv", header=T)
barplot(as.matrix(citysales[, 2:4]),
	  beside = T,
	  legend = citysales$City,
	  col = rainbow(5), # other color palette heat.colors()
	  border = "white",
	  ylab = "Revenue", cex.lab = 1.3,
	  las = 1)
```
---

## Miscellaneous graphs

* Correlation matrix

* Network graph

* Dynamic graph

---

## Correlation matrix plot
```{r, fig.align='center', echo=FALSE}
library(corrplot)
corrplot(cor(mtcars))
```
---

## Network graph
```{r, eval=T, echo=F, tidy=FALSE, fig.align='center'}
library(igraph)
library(gcookbook)
g <- graph.data.frame(madmen2, directed=T)
par(mar=c(0, 0 , 0, 0))
plot(g, layout=layout.fruchterman.reingold, vertex.size=9, edge.arrow.size=0.5,
     vertex.label=NA)
```
---

## Dynamic graphics
```{r, echo=F, results="asis", message=F}
library(googleVis)
m1 <- gvisMotionChart(Fruits, idvar="Fruit", timevar="Year")
print(m1, tag="chart")
```
---

## Conclusions

* R operates on data structure (and the fundamental type is vector)

* R manipulates data by functions, and R controls functions through
  parameters (both statistical functions and graphics function apply to the same principle)
  
---










